# Efficiency

```{r}
source(here::here("_chapter-setup.R"))
```

<!-- First 2025 working draft: copied from the 2024 chapter wording, then rebuilt from local PBWG daily taxi-time files. Taxi-in/out now includes Europe and the harmonised Brazil extract. ASMA is intentionally parked for this pass. -->

Operational efficiency is a critical component in assessing the management and execution of operations. It provides insights in the management of arrival and departure flows and the associated separation and synchronisation activities.
Inefficiencies can have an impact on user operations in terms of delays or excessive fuel burn.
In light of the previous chapters, it is therefore interesting to study how the available capacity was utilised to service demand during the different flight phases.

The measures reported in this comparison report are based on the observed travel time for surface operations (i.e. taxi-in and taxi-out) and during the arrival phase.
These travel times are compared with an associated reference time for a group of flights showing similar operational characteristics.
The determined difference (i.e. additional time) measures the level of inefficiency.
It must be noted that high performance operations will still yield a certain share of measured additional times.
Operational efficiency is therefore aiming at minimising rather than eliminating these additional times as they cannot be zero.

```{r}
txxt_years <- 2023:this_year
txxt_regions <- c("BRA", "EUR")
airport_names <- dplyr::bind_rows(
  bra_apts_names |> dplyr::mutate(REG = "BRA"),
  eur_apts_names |> dplyr::mutate(REG = "EUR")
)

read_txxt_year <- function(.region, .year) {
  readr::read_csv(
    here::here("data", glue::glue("PBWG-{.region}-txxt-analytic-{.year}-ref2024-icao_ganp_p20.csv")),
    show_col_types = FALSE
  ) |>
    dplyr::mutate(
      REG = .region,
      DATE = lubridate::ymd(DATE),
      YEAR = lubridate::year(DATE),
      AVG_TXXT = dplyr::if_else(MVTS_VALID > 0, TOT_TXXT / MVTS_VALID, NA_real_),
      AVG_ADD_TIME = dplyr::if_else(MVTS_VALID > 0, TOT_ADD_TIME / MVTS_VALID, NA_real_)
    )
}

txxt_daily <- tidyr::expand_grid(REG = txxt_regions, YEAR = txxt_years) |>
  purrr::pmap_dfr(\(REG, YEAR) read_txxt_year(REG, YEAR)) |>
  dplyr::filter(
    (REG == "BRA" & ICAO %in% bra_apts) |
      (REG == "EUR" & ICAO %in% eur_apts)
  )

txxt_annual <- txxt_daily |>
  dplyr::summarise(
    MVTS_VALID = sum(MVTS_VALID, na.rm = TRUE),
    MVTS_NA = sum(MVTS_NA, na.rm = TRUE),
    TOT_TXXT = sum(TOT_TXXT, na.rm = TRUE),
    TOT_REF = sum(TOT_REF, na.rm = TRUE),
    TOT_ADD_TIME = sum(TOT_ADD_TIME, na.rm = TRUE),
    AVG_TXXT = TOT_TXXT / MVTS_VALID,
    AVG_ADD_TIME = TOT_ADD_TIME / MVTS_VALID,
    .by = c(REG, ICAO, PHASE, YEAR)
  )

txxt_monthly <- txxt_daily |>
  dplyr::mutate(MONTH = lubridate::floor_date(DATE, "month")) |>
  dplyr::summarise(
    MVTS_VALID = sum(MVTS_VALID, na.rm = TRUE),
    MVTS_NA = sum(MVTS_NA, na.rm = TRUE),
    TOT_TXXT = sum(TOT_TXXT, na.rm = TRUE),
    TOT_REF = sum(TOT_REF, na.rm = TRUE),
    TOT_ADD_TIME = sum(TOT_ADD_TIME, na.rm = TRUE),
    AVG_TXXT = TOT_TXXT / MVTS_VALID,
    AVG_ADD_TIME = TOT_ADD_TIME / MVTS_VALID,
    .by = c(REG, ICAO, PHASE, MONTH)
  )

label_airports <- function(.data, .airport_col = ICAO) {
  .data |>
    dplyr::inner_join(airport_names, by = dplyr::join_by(REG, {{ .airport_col }} == ICAO)) |>
    dplyr::mutate(APT_LABEL = paste0({{ .airport_col }}, "\n", NAME))
}

plot_annual_txxt <- function(.phase, .region, .ylabel) {
  txxt_annual |>
    dplyr::filter(PHASE == .phase, REG == .region, YEAR %in% txxt_years) |>
    label_airports() |>
    dplyr::mutate(YEAR = as.factor(YEAR)) |>
    ggplot2::ggplot(ggplot2::aes(x = AVG_ADD_TIME, y = APT_LABEL, fill = YEAR)) +
    ggplot2::geom_col(position = ggplot2::position_dodge(-0.9), width = 0.9) +
    ggplot2::geom_vline(xintercept = c(2, 4), linetype = "dotted") +
    ggplot2::scale_fill_brewer(palette = "GnBu") +
    ggplot2::scale_x_continuous(limits = c(0, NA)) +
    ggplot2::labs(x = .ylabel, y = NULL, fill = NULL) +
    ggplot2::theme(
      legend.position = "top",
      legend.text = ggplot2::element_text(size = 8),
      legend.key.size = ggplot2::unit(0.3, "cm")
    )
}

plot_monthly_txxt <- function(.phase, .region, .ylabel, .ylim = NULL) {
  region_col <- getElement(bra_eur_colours, .region)
  p <- txxt_monthly |>
    dplyr::filter(PHASE == .phase, REG == .region) |>
    label_airports() |>
    ggplot2::ggplot(ggplot2::aes(x = MONTH, y = AVG_ADD_TIME, group = ICAO)) +
    ggplot2::geom_line(colour = region_col, linewidth = 0.35) +
    ggplot2::facet_wrap(. ~ APT_LABEL, ncol = 3) +
    ggplot2::labs(x = NULL, y = .ylabel) +
    ggplot2::theme(panel.spacing = ggplot2::unit(0.45, "cm"))
  if (!is.null(.ylim)) p <- p + ggplot2::coord_cartesian(ylim = .ylim)
  p
}

plot_txxt_mapping <- function() {
  txxt_annual |>
    dplyr::filter(YEAR %in% c(2024, 2025)) |>
    dplyr::mutate(PHASE = dplyr::recode(PHASE, ARR = "TXIT", DEP = "TXOT")) |>
    dplyr::select(REG, ICAO, PHASE, YEAR, AVG_ADD_TIME) |>
    tidyr::pivot_wider(names_from = PHASE, values_from = AVG_ADD_TIME) |>
    label_airports() |>
    dplyr::mutate(
      YEAR = as.character(YEAR),
      LABEL = dplyr::if_else(YEAR == "2025", ICAO, NA_character_)
    ) |>
    ggplot2::ggplot(ggplot2::aes(x = TXOT, y = TXIT, group = ICAO)) +
    ggplot2::geom_abline(intercept = 0, slope = 1, linetype = "dashed", colour = "grey60") +
    ggplot2::geom_path(colour = "lightblue") +
    ggplot2::geom_point(ggplot2::aes(shape = YEAR, colour = YEAR), size = 2) +
    ggplot2::facet_wrap(. ~ REG) +
    ggplot2::scale_colour_manual(values = c("lightblue", "blue")) +
    ggrepel::geom_label_repel(
      ggplot2::aes(label = LABEL),
      force = 100,
      max.overlaps = Inf,
      box.padding = 0.15,
      na.rm = TRUE,
      colour = "grey70",
      segment.colour = "grey70",
      size = 3
    ) +
    ggplot2::scale_x_continuous(limits = c(0, NA)) +
    ggplot2::scale_y_continuous(limits = c(0, NA)) +
    ggplot2::guides(shape = ggplot2::guide_legend(""), colour = ggplot2::guide_legend("")) +
    ggplot2::labs(
      x = "average additional taxi-out time [min/dep]",
      y = "average additional taxi-in time [min/arr]"
    ) +
    ggplot2::theme(
      legend.position = "top",
      legend.text = ggplot2::element_text(size = 8),
      legend.key.size = ggplot2::unit(0.3, "cm")
    )
}
```

## Additional Taxi-In Time

The additional taxi-in time measures the travel time of an arriving aircraft from its touchdown, i.e. the actual landing time, to its stand/gate position, i.e. actual in-block time).
This elapsed taxi-in time is compared to an anticipated reference time for aircraft arriving at the same runway and taxiing to the same (group of) stand/gate position(s).
Research showed that the taxi-times are not dependent on the type of aircraft.
The additional taxi-in time indicator provides a measure of the management of inbound surface traffic.

For this first 2025 update, the European taxi-in results are calculated from the PBWG APDF extraction.
The Brazilian taxi-in results use the harmonised DECEA taxi extract documented in `BRA-ingestion.qmd`.
For both regions, the reference time is determined from the 2024 sample using the ICAO GANP 20th percentile method by airport, phase, runway, and stand/gate position.
The analytic files retain the total observed taxi time (`TOT_TXXT`) across movements that are valid for the metric (`MVTS_VALID`), while `MVTS_NA` captures movements that could not be used because the taxi-time inputs were incomplete or no reference time was available.

### Annual Evolution of Additional Taxi-in Times

```{r}
#| label: fig-eur-txit-annual
#| fig-cap: Europe - Additional taxi-in time [min/arr] (2023-2025, 2024 GANP p20 reference)
#| fig-height: 4.8

plot_annual_txxt("ARR", "EUR", "average additional taxi-in time [min/arr]")
```

```{r}
#| label: fig-bra-txit-annual
#| fig-cap: Brazil - Additional taxi-in time [min/arr] (2023-2025, 2024 GANP p20 reference)
#| fig-height: 4.8

plot_annual_txxt("ARR", "BRA", "average additional taxi-in time [min/arr]")
```

The annual development of the average additional taxi-in times at the study airports is depicted by @fig-eur-txit-annual and @fig-bra-txit-annual.
The 2-minute threshold per arrival continues to serve as a practical reference point for evaluating taxi-in efficiency.
The 2025 update expands the European airport set to 12 airports, adding Athens (LGAV) and Istanbul (LTFM) to the comparison scope used in `_chapter-setup.R`.

### Monthly Variation of Additional Taxi-in Times

```{r}
#| label: fig-eur-txit-timeline
#| fig-cap: Europe - Monthly evolution of average additional taxi-in time
#| fig-height: 7.2

plot_monthly_txxt("ARR", "EUR", "avg. add. taxi-in time [min/arr]", .ylim = c(0, 8))
```

```{r}
#| label: fig-bra-txit-timeline
#| fig-cap: Brazil - Monthly evolution of average additional taxi-in time
#| fig-height: 7.2

plot_monthly_txxt("ARR", "BRA", "avg. add. taxi-in time [min/arr]", .ylim = c(0, 8))
```

The evolution of the taxi-in time at the study airports, shown in @fig-eur-txit-timeline and @fig-bra-txit-timeline, reinforces the annual comparison and provides a check on the seasonal variation of the surface movement indicator.
The interpretation remains deliberately light in this first draft until the harmonised Brazilian extract has been reviewed with DECEA.

## Taxi-Out Times

### Annual Evolution of Additional Taxi-out Times

```{r}
#| label: fig-eur-txot-annual
#| fig-cap: Europe - Average additional taxi-out time [min/dep] (2023-2025, 2024 GANP p20 reference)
#| fig-height: 4.8

plot_annual_txxt("DEP", "EUR", "average additional taxi-out time [min/dep]")
```

```{r}
#| label: fig-bra-txot-annual
#| fig-cap: Brazil - Average additional taxi-out time [min/dep] (2023-2025, 2024 GANP p20 reference)
#| fig-height: 4.8

plot_annual_txxt("DEP", "BRA", "average additional taxi-out time [min/dep]")
```

On average, higher additional times for taxi-out are observed across airports (c.f. @fig-eur-txot-annual and @fig-bra-txot-annual).
The taxi-out phase requires a higher level of management of the surface movements.
Departure procedures and flow control can impact the departure queue.

### Monthly Variation of Additional Taxi-out Times

```{r}
#| label: fig-eur-txot-timeline
#| fig-cap: Europe - Monthly evolution of average additional taxi-out time
#| fig-height: 7.2

plot_monthly_txxt("DEP", "EUR", "avg. add. taxi-out time [min/dep]", .ylim = c(0, 12))
```

```{r}
#| label: fig-bra-txot-timeline
#| fig-cap: Brazil - Monthly evolution of average additional taxi-out time
#| fig-height: 7.2

plot_monthly_txxt("DEP", "BRA", "avg. add. taxi-out time [min/dep]", .ylim = c(0, 12))
```

The monthly evolution in @fig-eur-txot-timeline and @fig-bra-txot-timeline provides the working view for updating the narrative once the Brazilian surface movement data are confirmed on the same basis.

## Mapping Additional Taxi-in and Taxi-out Times

```{r}
#| label: fig-txot-txit-mapping
#| fig-cap: Mapping of additional taxi-in and taxi-out times (2024-2025)
#| fig-height: 4.8

plot_txxt_mapping()
```

This analysis builds on the previous sections.
@fig-txot-txit-mapping compares the relationship between the taxi-in and taxi-out performance observed at the study airports.
It also shows that on average taxi-out operations accrue more additional time than taxi-in operations.

## Additional Time in Terminal Airspace

The additional time in terminal airspace is calculated as the difference of the actual flying time from entering the sequencing area to the actual landing time.
Previous research and guidance suggest that reference time can be built for flights sharing similar operational characteristics.

For this first 2025 run, the ASMA section is intentionally excluded because the sector boundaries still need to be agreed.

## Summary

This chapter analysed operational efficiency through the assessment of additional taxi-in and taxi-out times.
For Europe, the first 2025 update now uses the PBWG APDF extraction with a 2024 ICAO GANP 20th percentile reference and covers the 12 airports defined in `_chapter-setup.R`.
For Brazil, the chapter now uses the harmonised DECEA taxi extract prepared in APDF-like form and processed through the same PBWG taxi-time pipeline.
The chapter uses `TOT_TXXT` as the observed taxi-time total for the metric-ready sample and derives the additional-time averages from `MVTS_VALID`.
The next update should review the Brazilian harmonisation diagnostics and then move to the ASMA/DSMA sectorisation.
