---
title: "HFE - report proposal"
---

```{r}
#| label: setup
#| message: false
library(tidyverse)

ggplot2::theme_set(ggplot2::theme_minimal())
```



## Data preparation

### Brazil

DECEA -> data lake, HFE data for 2025
data ingestion via download from data lake

```{}
# setup ingestion
source("./R/rqutils-zip.R")

path_to_download <- "/Users/rainerkoelle/Downloads"
hfe_download <- fs::dir_ls(path_to_download, regexp = "OneDrive_1_29-04") |> basename()
hfe_fns      <- check_zip_content(path_to_download, hfe_download) |> pull(Name)

bra_kea <- read_zip_content(path_to_download, hfe_download, hfe_fns[1], "csv_auto")
bra_kep <- read_zip_content(path_to_download, hfe_download, hfe_fns[2], "csv_auto")

bra_kea <- bra_kea |> select(ADEP = plan_dep, ADES = real_arr, DATE_TIME = time, TYPE = equip
                  ,FLTID = id_icao, REG = reg
                  , HFE_BRA = kpi05_v1
                  ,DIST_ACHD = dist_achieved_v1
                  ,DIST_FLOWN = dist_flown_v1) |> 
  mutate(KPI = "KEA")

bra_kep <- bra_kep |> 
  select(ADEP = orig, ADES = dest, DATE_TIME = time   #, TYPE = equip
                  ,FLTID = id_icao
         #, REG = reg
         , HFE_BRA = kpi04_v1
                  ,DIST_ACHD = dist_achieved_v1
                  ,DIST_FLOWN = dist_flown_v1) |> 
  mutate(KPI = "KEP")
```

```{}
hfe_bra_network <- bind_rows(bra_kea, bra_kep) 
hfe_bra_network |> write_csv("./data/BRA-HFE-network-2025.csv")
```



### Europe

```{r}
hfe_eur_12 <- fs::dir_ls("./data", regexp = "EUR-hfe-Q") |> 
  purrr::map(.f = ~ arrow::read_parquet(.x)) |> bind_rows()

```

```{}
hfe_eur_network <- readr::read_csv("https://www.eurocontrol.int/performance/data/download/csv/horizontal_flight_efficiency_2025.csv", show_col_types = FALSE) |> 
  filter(ENTITY_NAME == "EUROCONTROL Area (PRR)")

hfe_eur_network |> write_csv("EUR-HFE-network-2025.csv")
```


### Airport Meta Data 

study scope and meta data

```{r}
bra_apts_names <- tibble::tribble(
  ~ICAO  , ~NAME
  ,"SBGR", "Guarulhos"
  ,"SBGL", "Galeão"
  ,"SBRJ", "Rio Dumont"
  ,"SBCF", "Belo Horizonte"
  ,"SBBR", "Brasília"
  ,"SBSV", "Salvador"
  ,"SBKP", "Campinas"
  ,"SBSP", "Congonhas"
  ,"SBCT", "Curitiba"
  ,"SBPA", "Porto Alegre"
  ,"SBRF", "Recife"
  ,"SBEG", "Eduardo Gomes"
)
bra_apts <- bra_apts_names$ICAO

eur_apts_names <- tibble::tribble(
  ~ICAO  , ~NAME
  ,"EGLL", "Heathrow"
  ,"EGKK", "Gatwick"
  ,"EHAM", "Amsterdam"
  ,"EDDF", "Frankfurt"
  ,"EDDM", "Munich"
  ,"LSZH", "Zurich"
  ,"LIRF", "Rome"
  ,"LFPG", "Paris"
  ,"LEMD", "Madrid"
  ,"LEBL", "Barcelona"
  ,"LPPT", "Lisbon"
  ,"LGAV", "Athens"
  ,"LTFM", "Istanbul"
)
eur_apts <- eur_apts_names$ICAO
```

get GCD

```{}
oa <- read_csv("https://davidmegginson.github.io/ourairports-data/airports.csv", show_col_types = FALSE)

bra_apts_latlon <- oa |> 
  filter(grepl("[A-Z]{4}", icao_code)) |> 
  select(ICAO = icao_code, LAT = latitude_deg, LON = longitude_deg) |> 
  filter(ICAO %in% bra_apts)

eur_apts_latlon <- oa |> 
  filter(grepl("[A-Z]{4}", icao_code)) |> 
  select(ICAO = icao_code, LAT = latitude_deg, LON = longitude_deg) |> 
  filter(ICAO %in% eur_apts)

bra_gcds <- expand_grid(ADEP = bra_apts, ADES = bra_apts) |> 
  left_join(bra_apts_latlon, by = join_by("ADEP" == "ICAO")) |> 
  rename(LAT1 = LAT, LON1 = LON) |> 
  left_join(bra_apts_latlon, by = join_by("ADES" == "ICAO")) |> 
  mutate( GCD = geosphere::distHaversine(cbind(LON1, LAT1), cbind(LON, LAT))
         ,GCD = GCD / 1852    # GCD in NM
         )

eur_gcds <- expand_grid(ADEP = eur_apts, ADES = eur_apts) |> 
    left_join(eur_apts_latlon, by = join_by("ADEP" == "ICAO")) |> rename(LAT1 = LAT, LON1 = LON) |> 
    left_join(eur_apts_latlon, by = join_by("ADES" == "ICAO")) |> 
    mutate( GCD = geosphere::distHaversine(cbind(LON1, LAT1), cbind(LON, LAT))
           ,GCD = GCD / 1852)

gcds <- bind_rows(bra_gcds |> mutate(REGION = "BRA"), eur_gcds |> mutate(REGION = "EUR")) |> 
  write_csv("./data/meta-apts-study-gcds.csv")
```

```{r}
gcds <- read_csv("./data/meta-apts-study-gcds.csv", show_col_types = F)

gcds |> 
ggplot() +
  geom_point(aes(x = GCD, y = REGION, color = REGION))
```


### Store Airport Pair Data for Study

filter to airport pairs

```{}
bra_kea <- bra_kea |> filter(ADEP %in% bra_apts & ADES %in% bra_apts)
bra_kep <- bra_kep |> filter(ADEP %in% bra_apts & ADES %in% bra_apts)

bind_rows(bra_kea, bra_kep) |> arrow::write_parquet("./data/BRA-HFE-12apts-2025.parquet")
```


## Analysis

```{r}
gcds <- read_csv("./data/meta-apts-study-gcds.csv", show_col_types = F)
```


```{r}
hfe_bra_12 <- arrow::read_parquet("./data/BRA-HFE-12apts-2025.parquet") |> 
  mutate(HFE = (DIST_FLOWN / DIST_ACHD) - 1, .after = HFE_BRA )

hfe_eur_12 <- fs::dir_ls("./data", regexp = "EUR-hfe-Q") |> 
  purrr::map(.f = ~ arrow::read_parquet(.x)) |> bind_rows()
```




top connections

```{r}
hfe_bra_stats <- hfe_bra_12 |> 
  group_by(ADEP, ADES) |> reframe(N = n()) |> 
  left_join(tmp |> select(ADEP, ADES, GCD), by = join_by(ADEP, ADES))
  
hfe_bra_stats |> arrange(desc(N))
```

```{r}
hfe_eur_stats <- hfe_eur_12 |> 
  group_by(ADEP, ADES) |> reframe(N = n()) |> 
  left_join(tmp |> select(ADEP, ADES, GCD), by = join_by(ADEP, ADES))
  
hfe_eur_stats |> arrange(desc(N))
```


library(plotly)

```{r}
tmp <- bind_rows(
  hfe_bra_stats |> mutate(REGION = "BRA", LABEL = paste0(ADEP,"-",ADES))
, hfe_eur_stats |> mutate(REGION = "EUR", LABEL = paste0(ADEP,"-",ADES))
) 

tmp |>  
  ggplot(aes(x = N, y = GCD, color = REGION, label = LABEL)) + 
  geom_point() 

ggplotly(p)
```

big-range
LPPT-LGAV
LTFM-LPPT
SBEG-SBGL
SBEG-SBGR

EHAM  LEMD  789.    LFPG  LPPT   3844  795
SBSV  SBGR  780
SBPA  SBGL  606

EGKK  LEBL   6587  600. EUR    EGKK-LEBL
 9 EGLL  LEBL   5786  620
LEBL  EDDM   4554  592.

SBCF  SBSP  12203  283. BRA    SBCF-SBSP
 2 LPPT  LEMD  11279  277. EUR    LPPT-LEMD
 3 LEMD  LPPT  11265  277. EUR    LEMD-LPPT
 4 SBSP  SBCF  10483  283. BRA    SBSP-SBCF
  LTFM  LGAV   5804  299. EUR    LTFM-LGAV
17 SBCF  SBKP   4987  269.


```{r}
be <- hfe_bra_12 |> mutate(DATE = date(DATE_TIME)) |> 
  group_by(DATE, ADEP, ADES, KPI) |> 
  reframe(N = n(), across(DIST_ACHD:DIST_FLOWN, ~ sum(.x))) |> 
  mutate(
    HFE = (DIST_FLOWN / DIST_ACHD) - 1, REGION = "BRA") |> 
  filter(HFE > 0, HFE < 0.25) |> 
  
  filter(ADEP == "SBEG", ADES == "SBGL")
  
  
  
 

rq <- hfe_eur_12 |> 
  mutate(DATE = date(ENTRY_TIME)) |> 
  group_by(DATE, ADEP, ADES, KPI) |> 
  reframe(N = n(), across(c("TOTAL_FLOWN_KM","TOTAL_ACHIEVED_KM"), ~ sum(.x))) |> 
  mutate(
    HFE = (TOTAL_FLOWN_KM / TOTAL_ACHIEVED_KM) - 1, REGION = "EUR") |> 
  
  filter(ADEP == "LPPT", ADES == "LGAV") 

```

```{r}
bind_rows(be, rq) |> 
  ggplot() +
  geom_path(aes(x = DATE, y = HFE, color = KPI, group = KPI)) +
  facet_wrap(vars(REGION))
```


## Network Level

```{r}
hfe_bra_network <- read_csv("./data/BRA-HFE-network-2025.csv", show_col_types = F) |> 
  mutate(
     DATE = lubridate::date(DATE_TIME)
    ,HFE = (DIST_FLOWN / DIST_ACHD) - 1, .after = HFE_BRA ) |> 
  filter(HFE > 0) |> 
  group_by(DATE, KPI) |> 
  reframe(N = n(), across(HFE:DIST_FLOWN, ~ sum(.x))) |> 
  mutate(HFE2 = (DIST_FLOWN / DIST_ACHD) - 1, REGION = "BRA") |> 
  select(REGION, DATE, HFE = HFE2, TYPE = KPI) |> 
  mutate(HFE = 0.5 * HFE)

hfe_eur_network <- read_csv("./data/EUR-HFE-network-2025.csv", show_col_types = FALSE) |> 
  filter(TYPE_MODEL %in% c("CPF","FTFM")) |> 
  mutate(HFE = (DIST_FLOWN_KM / DIST_ACHIEVED_KM) - 1, REGION = "EUR"
         ,TYPE_MODEL = case_when(TYPE_MODEL == "CPF" ~ "KEA", TYPE_MODEL == "FTFM" ~ "KEP", .default = NA)) |> 
  select(REGION, DATE = ENTRY_DATE, TYPE = TYPE_MODEL, HFE)
  

hfe_network <- bind_rows(
   hfe_bra_network
  ,hfe_eur_network
)
```

```{r}
hfe_network |> 
  ggplot() + geom_path(aes(x = DATE, y = HFE, color = TYPE, group = TYPE)) +
  scale_y_continuous(limits = c(0, NA)) +
  facet_wrap(vars(REGION))
```





