---
title: "HFE - report proposal"
---

```{r}
#| label: setup
#| message: false
library(tidyverse)

ggplot2::theme_set(ggplot2::theme_minimal())
```



## Data preparation

### Brazil

DECEA -> data lake, HFE data for 2025
data ingestion via download from data lake

```{}
# setup ingestion
source("./R/rqutils-zip.R")

path_to_download <- "/Users/rainerkoelle/Downloads"
hfe_download <- fs::dir_ls(path_to_download, regexp = "OneDrive_1_29-04") |> basename()
hfe_fns      <- check_zip_content(path_to_download, hfe_download) |> pull(Name)

bra_kea <- read_zip_content(path_to_download, hfe_download, hfe_fns[1], "csv_auto")
bra_kep <- read_zip_content(path_to_download, hfe_download, hfe_fns[2], "csv_auto")

bra_kea <- bra_kea |> select(ADEP = plan_dep, ADES = real_arr, DATE_TIME = time, TYPE = equip
                  ,FLTID = id_icao, REG = reg
                  , HFE_BRA = kpi05_v1
                  ,DIST_ACHD = dist_achieved_v1
                  ,DIST_FLOWN = dist_flown_v1) |> 
  mutate(KPI = "KEA")

bra_kep <- bra_kep |> 
  select(ADEP = orig, ADES = dest, DATE_TIME = time   #, TYPE = equip
                  ,FLTID = id_icao
         #, REG = reg
         , HFE_BRA = kpi04_v1
                  ,DIST_ACHD = dist_achieved_v1
                  ,DIST_FLOWN = dist_flown_v1) |> 
  mutate(KPI = "KEP")
```

```{}
hfe_bra_network <- bind_rows(bra_kea, bra_kep) 
hfe_bra_network |> write_csv("./data/BRA-HFE-network-2025.csv")
```



### Europe

```{r}
hfe_eur_12 <- fs::dir_ls("./data", regexp = "EUR-hfe-Q") |> 
  purrr::map(.f = ~ arrow::read_parquet(.x)) |> bind_rows()

```

```{}
hfe_eur_network <- readr::read_csv("https://www.eurocontrol.int/performance/data/download/csv/horizontal_flight_efficiency_2025.csv", show_col_types = FALSE) |> 
  filter(ENTITY_NAME == "EUROCONTROL Area (PRR)")

hfe_eur_network |> write_csv("EUR-HFE-network-2025.csv")
```


### Airport Meta Data 

study scope and meta data

```{r}
bra_apts_names <- tibble::tribble(
  ~ICAO  , ~NAME
  ,"SBGR", "Guarulhos"
  ,"SBGL", "Galeão"
  ,"SBRJ", "Rio Dumont"
  ,"SBCF", "Belo Horizonte"
  ,"SBBR", "Brasília"
  ,"SBSV", "Salvador"
  ,"SBKP", "Campinas"
  ,"SBSP", "Congonhas"
  ,"SBCT", "Curitiba"
  ,"SBPA", "Porto Alegre"
  ,"SBRF", "Recife"
  ,"SBEG", "Eduardo Gomes"
)
bra_apts <- bra_apts_names$ICAO

eur_apts_names <- tibble::tribble(
  ~ICAO  , ~NAME
  ,"EGLL", "Heathrow"
  ,"EGKK", "Gatwick"
  ,"EHAM", "Amsterdam"
  ,"EDDF", "Frankfurt"
  ,"EDDM", "Munich"
  ,"LSZH", "Zurich"
  ,"LFPG", "Paris"
  ,"LEMD", "Madrid"
  ,"LEBL", "Barcelona"
  ,"LPPT", "Lisbon"
  ,"LGAV", "Athens"
  ,"LTFM", "Istanbul"
)
eur_apts <- eur_apts_names$ICAO
```

get GCD

```{}
oa <- read_csv("https://davidmegginson.github.io/ourairports-data/airports.csv", show_col_types = FALSE)

bra_apts_latlon <- oa |> 
  filter(grepl("[A-Z]{4}", icao_code)) |> 
  select(ICAO = icao_code, LAT = latitude_deg, LON = longitude_deg) |> 
  filter(ICAO %in% bra_apts)

eur_apts_latlon <- oa |> 
  filter(grepl("[A-Z]{4}", icao_code)) |> 
  select(ICAO = icao_code, LAT = latitude_deg, LON = longitude_deg) |> 
  filter(ICAO %in% eur_apts)

bra_gcds <- expand_grid(ADEP = bra_apts, ADES = bra_apts) |> 
  left_join(bra_apts_latlon, by = join_by("ADEP" == "ICAO")) |> 
  rename(LAT1 = LAT, LON1 = LON) |> 
  left_join(bra_apts_latlon, by = join_by("ADES" == "ICAO")) |> 
  mutate( GCD = geosphere::distHaversine(cbind(LON1, LAT1), cbind(LON, LAT))
         ,GCD = GCD / 1852    # GCD in NM
         )

eur_gcds <- expand_grid(ADEP = eur_apts, ADES = eur_apts) |> 
    left_join(eur_apts_latlon, by = join_by("ADEP" == "ICAO")) |> rename(LAT1 = LAT, LON1 = LON) |> 
    left_join(eur_apts_latlon, by = join_by("ADES" == "ICAO")) |> 
    mutate( GCD = geosphere::distHaversine(cbind(LON1, LAT1), cbind(LON, LAT))
           ,GCD = GCD / 1852)

gcds <- bind_rows(bra_gcds |> mutate(REGION = "BRA"), eur_gcds |> mutate(REGION = "EUR")) |> 
  write_csv("./data/meta-apts-study-gcds.csv")
```

```{r}
gcds <- read_csv("./data/meta-apts-study-gcds.csv", show_col_types = F)

gcds |> 
ggplot() +
  geom_point(aes(x = GCD, y = REGION, color = REGION))
```


### Store Airport Pair Data for Study

filter to airport pairs

```{}
bra_kea <- bra_kea |> filter(ADEP %in% bra_apts & ADES %in% bra_apts)
bra_kep <- bra_kep |> filter(ADEP %in% bra_apts & ADES %in% bra_apts)

bind_rows(bra_kea, bra_kep) |> arrow::write_parquet("./data/BRA-HFE-12apts-2025.parquet")
```


## Analysis

The following chunks use the prepared data above.  Brazilian KPI values are not
used directly; HFE is recalculated from achieved/flown distances and infeasible
negative values are retained only in quality counters.

```{r}
HFE_VALID_MAX <- 0.25

gcds <- read_csv("./data/meta-apts-study-gcds.csv", show_col_types = F)
```


```{r}
hfe_bra_12 <- arrow::read_parquet("./data/BRA-HFE-12apts-2025.parquet") |> 
  transmute(
    ADEP, ADES, DATE_TIME, FLTID, REG, HFE_BRA,
    HFE_TYPE = KPI,
    DIST_FLOWN,
    DIST_ACHIEVED = DIST_ACHD,
    HFE = (DIST_FLOWN / DIST_ACHD) - 1
  )

hfe_eur_12 <- fs::dir_ls("./data", regexp = "EUR-hfe-Q") |> 
  purrr::map(.f = ~ arrow::read_parquet(.x)) |> bind_rows() |> 
  filter(ADEP %in% eur_apts, ADES %in% eur_apts) |> 
  transmute(
    ADEP, ADES, DATE_TIME = ENTRY_TIME, HFE_TYPE = KPI,
    DIST_FLOWN = TOTAL_FLOWN_KM / 1.852,
    DIST_ACHIEVED = TOTAL_ACHIEVED_KM / 1.852,
    HFE = (TOTAL_FLOWN_KM / TOTAL_ACHIEVED_KM) - 1
  )
```




top connections

```{r}
hfe_bra_stats <- hfe_bra_12 |> 
  group_by(ADEP, ADES) |> 
  reframe(
    N = n(),
    N_VALID = sum(is.finite(HFE) & HFE >= 0 & HFE <= HFE_VALID_MAX),
    N_NEG = sum(HFE < 0, na.rm = TRUE),
    N_HIGH = sum(HFE > HFE_VALID_MAX, na.rm = TRUE)
  ) |> 
  left_join(gcds |> filter(REGION == "BRA") |> select(ADEP, ADES, GCD), by = join_by(ADEP, ADES))
  
hfe_bra_stats |> arrange(desc(N))
```

```{r}
hfe_eur_stats <- hfe_eur_12 |> 
  group_by(ADEP, ADES) |> 
  reframe(
    N = n(),
    N_VALID = sum(is.finite(HFE) & HFE >= 0 & HFE <= HFE_VALID_MAX),
    N_NEG = sum(HFE < 0, na.rm = TRUE),
    N_HIGH = sum(HFE > HFE_VALID_MAX, na.rm = TRUE)
  ) |> 
  left_join(gcds |> filter(REGION == "EUR") |> select(ADEP, ADES, GCD), by = join_by(ADEP, ADES))
  
hfe_eur_stats |> arrange(desc(N))
```


Interactive plot used to identify airport pairs with broadly similar great-circle
distance and enough flights.

```{r}
pair_screening <- bind_rows(
  hfe_bra_stats |> mutate(REGION = "BRA", LABEL = paste0(ADEP,"-",ADES))
, hfe_eur_stats |> mutate(REGION = "EUR", LABEL = paste0(ADEP,"-",ADES))
) 

pair_screening_plot <- pair_screening |>  
  ggplot(aes(x = N, y = GCD, color = REGION, label = LABEL)) + 
  geom_point()

if (requireNamespace("plotly", quietly = TRUE)) {
  plotly::ggplotly(pair_screening_plot)
} else {
  pair_screening_plot
}
```

Candidate pairs retained for the report work-up.  The one-way selections from
the screening plot are expanded to both directions for the analytic dataset.

```{r}
candidate_pairs_one_way <- tribble(
  ~PAIR_GROUP, ~ADEP, ~ADES,
  "long",      "LPPT", "LGAV",
  "long",      "LTFM", "LPPT",
  "long",      "SBEG", "SBGL",
  "long",      "SBEG", "SBGR",
  "mid_600",   "SBBR", "SBSV",
  "mid_600",   "SBBR", "SBCT",
  "mid_600",   "SBPA", "SBGL",
  "mid_600",   "EGKK", "LEBL",
  "mid_600",   "EGLL", "LEBL",
  "mid_600",   "LEBL", "EDDM",
  "mid_780_800", "EHAM", "LEMD",
  "mid_780_800", "LFPG", "LPPT",
  "mid_780_800", "SBSV", "SBGR",
  "mid_780_800", "SBSV", "SBKP",
  "mid_780_800", "SBSV", "SBSP",
  "short",     "SBCF", "SBSP",
  "short",     "LPPT", "LEMD",
  "short",     "LTFM", "LGAV",
  "short",     "SBCF", "SBKP"
)

candidate_pairs <- bind_rows(
  candidate_pairs_one_way,
  candidate_pairs_one_way |> rename(ADEP_REV = ADES, ADES_REV = ADEP) |> transmute(PAIR_GROUP, ADEP = ADEP_REV, ADES = ADES_REV)
) |> 
  distinct() |> 
  left_join(
    gcds |> select(REGION, ADEP, ADES, GCD),
    by = join_by(ADEP, ADES)
  )

candidate_pairs |> arrange(PAIR_GROUP, REGION, GCD, ADEP, ADES)

```

```{r}
pair_example <- bind_rows(
  hfe_bra_12 |> mutate(REGION = "BRA"),
  hfe_eur_12 |> mutate(REGION = "EUR")
) |> 
  semi_join(candidate_pairs, by = join_by(ADEP, ADES)) |> 
  filter((ADEP == "SBEG" & ADES == "SBGL") | (ADEP == "LPPT" & ADES == "LGAV")) |> 
  mutate(DATE = date(DATE_TIME)) |> 
  group_by(REGION, DATE, ADEP, ADES, HFE_TYPE) |> 
  reframe(
    N = n(),
    N_VALID = sum(is.finite(HFE) & HFE >= 0 & HFE <= HFE_VALID_MAX),
    N_NEG = sum(HFE < 0, na.rm = TRUE),
    N_HIGH = sum(HFE > HFE_VALID_MAX, na.rm = TRUE),
    DIST_FLOWN = sum(DIST_FLOWN[is.finite(HFE) & HFE >= 0 & HFE <= HFE_VALID_MAX], na.rm = TRUE),
    DIST_ACHIEVED = sum(DIST_ACHIEVED[is.finite(HFE) & HFE >= 0 & HFE <= HFE_VALID_MAX], na.rm = TRUE)
  ) |> 
  mutate(HFE = if_else(DIST_ACHIEVED > 0, (DIST_FLOWN / DIST_ACHIEVED) - 1, NA_real_))

pair_example |> 
  ggplot() +
  geom_path(aes(x = DATE, y = HFE, color = HFE_TYPE, group = HFE_TYPE)) +
  facet_wrap(vars(REGION))
```


## Network Level

```{r}
hfe_bra_network <- read_csv("./data/BRA-HFE-network-2025.csv", show_col_types = F) |> 
  transmute(
    DATE = lubridate::date(DATE_TIME),
    HFE_TYPE = KPI,
    DIST_FLOWN,
    DIST_ACHIEVED = DIST_ACHD,
    HFE = (DIST_FLOWN / DIST_ACHD) - 1,
    IS_VALID = is.finite(HFE) & HFE >= 0 & HFE <= HFE_VALID_MAX
  ) |> 
  group_by(DATE, HFE_TYPE) |> 
  reframe(
    N = n(),
    N_VALID = sum(IS_VALID),
    N_NEG = sum(HFE < 0, na.rm = TRUE),
    N_HIGH = sum(HFE > HFE_VALID_MAX, na.rm = TRUE),
    DIST_FLOWN = sum(DIST_FLOWN[IS_VALID], na.rm = TRUE),
    DIST_ACHIEVED = sum(DIST_ACHIEVED[IS_VALID], na.rm = TRUE)
  ) |> 
  mutate(HFE = if_else(DIST_ACHIEVED > 0, (DIST_FLOWN / DIST_ACHIEVED) - 1, NA_real_), REGION = "BRA")

hfe_eur_network <- read_csv("./data/EUR-HFE-network-2025.csv", show_col_types = FALSE) |> 
  filter(TYPE_MODEL %in% c("CPF","FTFM")) |> 
  transmute(
    REGION = "EUR",
    DATE = lubridate::ymd(ENTRY_DATE),
    HFE_TYPE = case_when(TYPE_MODEL == "CPF" ~ "KEA", TYPE_MODEL == "FTFM" ~ "KEP", .default = NA),
    N = NA_integer_,
    N_VALID = NA_integer_,
    N_NEG = NA_integer_,
    N_HIGH = NA_integer_,
    DIST_FLOWN = DIST_FLOWN_KM / 1.852,
    DIST_ACHIEVED = DIST_ACHIEVED_KM / 1.852,
    HFE = (DIST_FLOWN / DIST_ACHIEVED) - 1
  )
  

hfe_network <- bind_rows(
   hfe_bra_network
  ,hfe_eur_network
) |> 
  select(REGION, DATE, HFE_TYPE, N, N_VALID, N_NEG, N_HIGH, DIST_FLOWN, DIST_ACHIEVED, HFE)
```

```{r}
hfe_network |> 
  ggplot() + geom_path(aes(x = DATE, y = HFE, color = HFE_TYPE, group = HFE_TYPE)) +
  scale_y_continuous(limits = c(0, NA)) +
  facet_wrap(vars(REGION))
```

```{r}
readr::write_csv(hfe_network, "./data/hfe-network-daily-2025.csv")
```


## Airport Pair Level

```{r}
hfe_pair_daily <- bind_rows(
  hfe_bra_12 |> mutate(REGION = "BRA"),
  hfe_eur_12 |> mutate(REGION = "EUR")
) |> 
  semi_join(candidate_pairs, by = join_by(ADEP, ADES)) |> 
  mutate(
    DATE = lubridate::date(DATE_TIME),
    IS_VALID = is.finite(HFE) & HFE >= 0 & HFE <= HFE_VALID_MAX
  ) |> 
  group_by(REGION, DATE, ADEP, ADES, HFE_TYPE) |> 
  reframe(
    N = n(),
    N_VALID = sum(IS_VALID),
    N_NEG = sum(HFE < 0, na.rm = TRUE),
    N_HIGH = sum(HFE > HFE_VALID_MAX, na.rm = TRUE),
    DIST_FLOWN = sum(DIST_FLOWN[IS_VALID], na.rm = TRUE),
    DIST_ACHIEVED = sum(DIST_ACHIEVED[IS_VALID], na.rm = TRUE)
  ) |> 
  mutate(HFE = if_else(DIST_ACHIEVED > 0, (DIST_FLOWN / DIST_ACHIEVED) - 1, NA_real_)) |> 
  left_join(candidate_pairs, by = join_by(REGION, ADEP, ADES)) |> 
  select(REGION, DATE, PAIR_GROUP, ADEP, ADES, GCD, HFE_TYPE, N, N_VALID, N_NEG, N_HIGH, DIST_FLOWN, DIST_ACHIEVED, HFE)

hfe_pair_daily |> 
  count(REGION, PAIR_GROUP, ADEP, ADES, HFE_TYPE, wt = N_VALID, name = "N_VALID") |> 
  arrange(PAIR_GROUP, REGION, desc(N_VALID))
```

```{r}
hfe_pair_daily |> 
  ggplot(aes(x = DATE, y = HFE, color = HFE_TYPE, group = HFE_TYPE)) +
  geom_path() +
  facet_wrap(vars(REGION, PAIR_GROUP), scales = "free_y") +
  scale_y_continuous(limits = c(0, NA))
```

```{r}
readr::write_csv(hfe_pair_daily, "./data/hfe-airport-pairs-daily-2025.csv")
```
