---
title: "Brazil ASMA Ingestion Check"
editor: visual
format:
  html:
    toc: true
    toc-depth: 3
---

```{r}
#| label: setup
#| message: false
#| warning: false

source(here::here("_chapter-setup.R"))
library(data.table)
library(fs)
library(knitr)

annual_zip <- "/Users/rainerkoelle/Downloads/OneDrive_1_22-04-2026.zip"
monthly_zip <- "/Users/rainerkoelle/Downloads/OneDrive_2_22-04-2026.zip"

asma_archives <- tibble::tribble(
  ~DATASET, ~ZIP_PATH, ~NOTE,
  "annual_2023_2025", annual_zip, "Three annual files, semicolon-delimited.",
  "monthly_2025", monthly_zip, "Monthly 2025 files, comma-delimited, with a leading row index."
)

expected_apdf_fields <- c(
  "AP_C_FLTID",
  "ADEP_ICAO", "ADES_ICAO",
  "AP_C_REG", "ARCTYP", "AC_CLASS",
  "AP_C_RWY", "AP_C_STND",
  "MVT_TIME_UTC", "BLOCK_TIME_UTC", "SCHED_TIME_UTC",
  "C40_CROSS_TIME", "C40_BEARING",
  "C100_CROSS_TIME", "C100_BEARING",
  "SRC_PHASE",
  "IM_SAMAD_ID",
  "SRC_AIRPORT"
)

apdf_mapping <- tibble::tribble(
  ~APDF_FIELD, ~SOURCE_FIELD, ~STATUS, ~NOTE,
  "AP_C_FLTID", "fltid", "available", "Flight identifier.",
  "ADEP_ICAO", "adep", "available", "Departure aerodrome.",
  "ADES_ICAO", "ades", "available", "Arrival aerodrome.",
  "AP_C_REG", "reg", "available", "Aircraft registration is present but often blank.",
  "ARCTYP", "type", "available", "Aircraft type.",
  "AC_CLASS", "CLASS", "available", "Aircraft weight class.",
  "AP_C_RWY", "drwy_validado / drwy", "available", "Use validated runway when populated, otherwise raw runway.",
  "AP_C_STND", "stnd", "available", "Stand/gate field.",
  "MVT_TIME_UTC", "aldt", "available for ARR", "Landing time is the natural arrival movement time.",
  "BLOCK_TIME_UTC", "aibt", "available for ARR", "In-block time; blank where no match.",
  "SCHED_TIME_UTC", "sibt", "available for ARR", "Scheduled in-block time; often blank in the sample.",
  "C40_CROSS_TIME", NA_character_, "not named as APDF", "The new files provide one generic crossing time field, c_time.",
  "C40_BEARING", NA_character_, "not named as APDF", "The new files provide one generic bearing field, bear.",
  "C100_CROSS_TIME", NA_character_, "not named as APDF", "No explicit C100 field in the new files.",
  "C100_BEARING", NA_character_, "not named as APDF", "No explicit C100 bearing field in the new files.",
  "SRC_PHASE", NA_character_, "implicit ARR", "Files appear to be ASMA/arrival-oriented; no explicit phase column.",
  "IM_SAMAD_ID", "id", "available", "Source movement identifier.",
  "SRC_AIRPORT", "ades", "available for ARR", "For arrivals, the source airport is the destination aerodrome."
)

read_zip_head <- function(zip_path, member, sep, n = 50000) {
  data.table::fread(
    cmd = sprintf("unzip -p %s %s", shQuote(zip_path), shQuote(member)),
    sep = sep,
    nrows = n,
    na.strings = c("", "NA", " "),
    colClasses = "character",
    showProgress = FALSE,
    data.table = FALSE
  ) |>
    tibble::as_tibble()
}

normalise_names <- function(x) {
  x |>
    rename_with(~ sub("^V1$", "row_index", .x)) |>
    rename_with(~ gsub('^"|"$', "", .x))
}

read_asma_sample <- function(dataset, member, n = 50000) {
  archive <- dplyr::filter(asma_archives, DATASET == dataset)
  sep <- if (dataset == "annual_2023_2025") ";" else ","

  read_zip_head(archive$ZIP_PATH, member, sep = sep, n = n) |>
    normalise_names() |>
    mutate(
      DATASET = dataset,
      MEMBER = member,
      YEAR = stringr::str_extract(member, "20[0-9]{2}"),
      MONTH_TOKEN = stringr::str_match(member, "kpi08_([a-z]{3})20[0-9]{2}\\.csv$")[, 2],
      .before = 1
    )
}

present <- function(x) {
  !is.na(x) & trimws(as.character(x)) != ""
}

summarise_presence <- function(x) {
  n <- nrow(x)
  tibble::tibble(
    N_SAMPLE = n,
    ARR_LIKE_ADES_SB = sum(grepl("^SB", x$ades), na.rm = TRUE),
    DEP_LIKE_ADEP_SB = sum(grepl("^SB", x$adep), na.rm = TRUE),
    KPI08_POPULATED = sum(present(x$kpi08), na.rm = TRUE),
    TRANSITO_POPULATED = sum(present(x$transito), na.rm = TRUE),
    DESIMP_POPULATED = sum(present(x$desimp), na.rm = TRUE),
    C_TIME_POPULATED = sum(present(x$c_time), na.rm = TRUE),
    BEAR_POPULATED = sum(present(x$bear), na.rm = TRUE),
    AIBT_POPULATED = sum(present(x$aibt), na.rm = TRUE),
    ALDT_POPULATED = sum(present(x$aldt), na.rm = TRUE),
    SIBT_POPULATED = sum(present(x$sibt), na.rm = TRUE)
  ) |>
    mutate(across(-N_SAMPLE, ~ round(.x / N_SAMPLE, 3), .names = "{.col}_SHARE"))
}

parse_duration_minutes <- function(x) {
  x <- trimws(as.character(x))
  x[x == "" | x == "NA"] <- NA_character_
  sign <- ifelse(grepl("^-", x), -1, 1)
  x <- sub("^-", "", x)

  parts <- stringr::str_match(x, "^([0-9]+):([0-9]{2}):([0-9]{2}(?:\\.[0-9]+)?)$")
  out <- suppressWarnings(
    as.numeric(parts[, 2]) * 60 +
      as.numeric(parts[, 3]) +
      as.numeric(parts[, 4]) / 60
  )
  sign * out
}
```

This note is a first-pass ingestion check for the two new Brazil ASMA archives received on 22 April 2026.
It reads the zip files directly from Downloads and samples each source member, without unpacking the large CSVs into the project tree.

The checks below focus on:

-   coverage and file layout;
-   whether APDF-like input fields can be mapped;
-   whether `kpi08` and additional ASMA time fields are already calculated;
-   whether the payload is arrival-only or contains usable departure/DSMA evidence.

## Archive Inventory

```{r}
#| label: tbl-archive-inventory

inventory <- purrr::pmap_dfr(
  asma_archives,
  function(DATASET, ZIP_PATH, NOTE) {
    utils::unzip(ZIP_PATH, list = TRUE) |>
      tibble::as_tibble() |>
      transmute(
        DATASET = DATASET,
        MEMBER = Name,
        SIZE_MB = round(Length / 1024^2, 1),
        DATE = Date
      )
  }
)

inventory |>
  mutate(
    YEAR = stringr::str_extract(MEMBER, "20[0-9]{2}"),
    MONTH_TOKEN = stringr::str_match(MEMBER, "kpi08_([a-z]{3})20[0-9]{2}\\.csv$")[, 2]
  ) |>
  arrange(DATASET, MEMBER) |>
  kable()
```

The annual archive contains one file per year for 2023, 2024, and 2025.
The monthly archive contains 2025 monthly files.
It currently includes both `kpi08_feb2025.csv` and `kpi08_fev2025.csv`, which should be checked as a likely duplicate or Portuguese/English month-name export issue before combining all months.

## Field Inventory

```{r}
#| label: tbl-field-inventory

field_inventory <- inventory |>
  mutate(
    SEP = if_else(DATASET == "annual_2023_2025", ";", ","),
    ZIP_PATH = asma_archives$ZIP_PATH[match(DATASET, asma_archives$DATASET)],
    FIELDS = purrr::pmap(
      list(ZIP_PATH, MEMBER, SEP),
      function(zip_path, member, sep) {
        names(read_zip_head(
          zip_path = zip_path,
          member = member,
          sep = sep,
          n = 5
        ) |> normalise_names())
      }
    )
  ) |>
  tidyr::unnest_longer(FIELDS, values_to = "FIELD") |>
  count(DATASET, FIELD, name = "N_FILES") |>
  arrange(DATASET, FIELD)

field_inventory |>
  kable()
```

```{r}
#| label: tbl-field-differences

field_inventory |>
  select(DATASET, FIELD) |>
  mutate(PRESENT = TRUE) |>
  tidyr::pivot_wider(names_from = DATASET, values_from = PRESENT, values_fill = FALSE) |>
  arrange(FIELD) |>
  kable()
```

Aside from the leading `row_index` in the monthly archive and the delimiter difference, both new data sets expose the same operational fields: `id`, airport pair, flight id, registration, aircraft class/type, runway/stand, `bear`, `c_time`, `aibt`, `aldt`, `sibt`, `setor`, `transito`, `c`, `desimp`, and `kpi08`.

## APDF Readiness

```{r}
#| label: tbl-apdf-readiness

apdf_mapping |>
  mutate(APDF_EXPECTED = APDF_FIELD %in% expected_apdf_fields) |>
  select(APDF_FIELD, SOURCE_FIELD, STATUS, NOTE) |>
  kable()
```

The files are close to an APDF arrival extract, but not a drop-in APDF source:

-   the basic APDF identity, airport, aircraft, runway, stand, movement, block, and scheduled fields are available or can be mapped;
-   `C40_CROSS_TIME`/`C40_BEARING` and `C100_CROSS_TIME`/`C100_BEARING` are not explicitly named;
-   the supplied `c_time` and `bear` likely represent one ASMA crossing point, but the file itself does not say whether that point is C40, C100, or another ASMA reference;
-   there is no explicit `SRC_PHASE`; the data appear arrival-oriented because `ades` is the Brazilian airport and `aldt`/`aibt` are arrival timestamps.

## Sample Presence Checks

```{r}
#| label: sample-members

sample_members <- tibble::tribble(
  ~DATASET, ~MEMBER,
  "annual_2023_2025", "kpi08_2023.csv",
  "annual_2023_2025", "kpi08_2024.csv",
  "annual_2023_2025", "kpi08_2025.csv",
  "monthly_2025", "kpi08_jan2025.csv",
  "monthly_2025", "kpi08_feb2025.csv",
  "monthly_2025", "kpi08_fev2025.csv",
  "monthly_2025", "kpi08_dec2025.csv"
)

samples <- sample_members |>
  mutate(DATA = purrr::map2(DATASET, MEMBER, read_asma_sample, n = 50000))
```

```{r}
#| label: tbl-sample-presence

samples |>
  mutate(SUMMARY = purrr::map(DATA, summarise_presence)) |>
  select(DATASET, MEMBER, SUMMARY) |>
  tidyr::unnest(SUMMARY) |>
  select(
    DATASET, MEMBER, N_SAMPLE,
    ends_with("_SHARE")
  ) |>
  arrange(DATASET, MEMBER) |>
  kable()
```

Interpretation notes:

-   `KPI08_POPULATED_SHARE` confirms that `kpi08` is already calculated in both new data sets for records with sufficient input;
-   `TRANSITO_POPULATED_SHARE` and `DESIMP_POPULATED_SHARE` show that additional ASMA time fields are present;
-   `C_TIME_POPULATED_SHARE` and `BEAR_POPULATED_SHARE` indicate whether crossing-point evidence exists for reconstructing or validating the ASMA calculation;
-   `ARR_LIKE_ADES_SB_SHARE` versus `DEP_LIKE_ADEP_SB_SHARE` is a quick phase clue: a high `ades` Brazil share is consistent with arrivals into Brazil, while domestic flights naturally have both `adep` and `ades` in Brazil.

## KPI08 and ASMA-Time Distributions

```{r}
#| label: tbl-time-distributions

samples |>
  mutate(
    DIST = purrr::map(DATA, function(x) {
      x |>
        transmute(
          kpi08_min = parse_duration_minutes(kpi08),
          transito_min = parse_duration_minutes(transito),
          desimp_min = parse_duration_minutes(desimp)
        ) |>
        summarise(
          KPI08_N = sum(!is.na(kpi08_min)),
          KPI08_MED = median(kpi08_min, na.rm = TRUE),
          KPI08_P95 = quantile(kpi08_min, 0.95, na.rm = TRUE),
          TRANSITO_N = sum(!is.na(transito_min)),
          TRANSITO_MED = median(transito_min, na.rm = TRUE),
          TRANSITO_P95 = quantile(transito_min, 0.95, na.rm = TRUE),
          DESIMP_N = sum(!is.na(desimp_min)),
          DESIMP_MED = median(desimp_min, na.rm = TRUE),
          DESIMP_P95 = quantile(desimp_min, 0.95, na.rm = TRUE)
        )
    })
  ) |>
  select(DATASET, MEMBER, DIST) |>
  tidyr::unnest(DIST) |>
  mutate(across(where(is.numeric), ~ round(.x, 2))) |>
  kable()
```

`transito`, `desimp`, and `kpi08` parse as durations in the sample rows.
This is useful because the source appears to provide both the calculated KPI08 and enough intermediate ASMA timing to sanity-check the formula at least for populated rows.

## Arrival and Departure Clues

```{r}
#| label: tbl-airport-pair-clues

samples |>
  mutate(
    AIRPORT_PAIRS = purrr::map(DATA, function(x) {
      x |>
        mutate(
          ADEP_BRA = grepl("^SB", adep),
          ADES_BRA = grepl("^SB", ades),
          PAIR_TYPE = case_when(
            ADEP_BRA & ADES_BRA ~ "domestic_BRA",
            !ADEP_BRA & ADES_BRA ~ "international_arr_to_BRA",
            ADEP_BRA & !ADES_BRA ~ "international_dep_from_BRA",
            TRUE ~ "non_BRA_pair"
          )
        ) |>
        count(PAIR_TYPE, name = "N") |>
        mutate(SHARE = round(N / sum(N), 3))
    })
  ) |>
  select(DATASET, MEMBER, AIRPORT_PAIRS) |>
  tidyr::unnest(AIRPORT_PAIRS) |>
  arrange(DATASET, MEMBER, desc(N)) |>
  kable()
```

The new ASMA files do not look like departure DSMA files because they use arrival fields (`aldt`, `aibt`, `sibt`) and have no explicit departure movement or off-block/take-off timestamps.
The presence of Brazilian `adep` values mostly reflects domestic arrivals where both ends of the pair are in Brazil.

## Preliminary Conclusions

-   Both new archives provide calculated `kpi08`.
-   Both provide additional ASMA time fields: `transito` and `desimp`, plus `c_time` and `bear` for a crossing-point reference.
-   The main structural difference is annual versus monthly packaging, delimiter, and the monthly archive's leading row index.
-   The monthly archive needs a duplicate check for `feb` versus `fev`.
-   The data are arrival-oriented ASMA data and appear usable for ARR KPI08/APDF-like ingestion.
-   The files do not appear to provide true DEP/DSMA evidence; a separate DSMA source would still be needed unless DECEA confirms that departure fields are encoded elsewhere.

## Next Ingestion Steps

1.  Confirm with DECEA what `c_time`/`bear` represent: C40, C100, or another ASMA reference.
2.  Confirm whether `transito`, `desimp`, and `kpi08` definitions match KPI08 calculation expectations.
3.  Decide whether to use the annual 2025 file or reconstruct 2025 from monthly files after resolving `feb`/`fev`.
4.  Build a harmonised parquet extract with APDF-like names for ARR only, retaining the original calculated KPI fields for validation.

## Production Ingestion

The first production-style ingestion is implemented in `R/08-ingest-bra-asma.R`.
It uses the multi-year annual archive and keeps the raw Brazil `CLASS` value as supplied.
At this stage, `CLASS` should therefore be interpreted as the Brazil/WTC-style category available in the source, not as the APDF class code that combines wake-turbulence category with propulsion.

The script writes:

-   APDF-like ARR ASMA parquet extracts to `/Users/rainerkoelle/RProjects/PBWG-BRA/data/asma`;
-   2024 GANP p20 reference files to `/Users/rainerkoelle/RProjects/PBWG-BRA/outputs/asma-reference-2024`;
-   daily analytic ASMA files to `/Users/rainerkoelle/RProjects/PBWG-BRA/outputs/asma-daily-2023`, `asma-daily-2024`, and `asma-daily-2025`;
-   combined report-layer outputs to `data/`.

The reference is calculated from 2024 valid arrival ASMA samples by:

`ICAO`, `PHASE`, `RANGE`, `CLASS`, `RWY`, and `SECTOR_GROUP`.

The daily analytic output keeps the same breakdown fields and reports movement counts plus total observed ASMA time, total reference time, total additional time, and the Brazil-supplied reference/KPI08 totals for comparison.

```{r}
#| label: tbl-production-output-check

asma_daily_path <- here::here("data", "PBWG-BRA-asma-analytic-2023-2025-ref2024-icao_ganp_p20.csv")
asma_coverage_path <- here::here("data", "BRA-asma-coverage-summary-2023-2025.csv")
asma_reference_coverage_path <- here::here("data", "BRA-asma-reference-coverage-2024.csv")

if (file.exists(asma_daily_path)) {
  asma_daily <- readr::read_csv(asma_daily_path, show_col_types = FALSE)

  asma_daily |>
    summarise(
      DAILY_GROUPS = n(),
      MVTS_VALID = sum(MVTS_VALID, na.rm = TRUE),
      MVTS_NA = sum(MVTS_NA, na.rm = TRUE),
      TOT_ASMA = sum(TOT_ASMA, na.rm = TRUE),
      TOT_REF = sum(TOT_REF, na.rm = TRUE),
      TOT_ADD_TIME = sum(TOT_ADD_TIME, na.rm = TRUE),
      .by = c(PHASE, RANGE)
    ) |>
    mutate(across(where(is.numeric), ~ round(.x, 1))) |>
    kable()
} else {
  tibble::tibble(NOTE = "Run R/08-ingest-bra-asma.R to create the production output.") |>
    kable()
}
```

```{r}
#| label: tbl-production-coverage-check

if (file.exists(asma_coverage_path)) {
  readr::read_csv(asma_coverage_path, show_col_types = FALSE) |>
    summarise(
      N_TOTAL = sum(N_TOTAL, na.rm = TRUE),
      N_VALID_ASMA = sum(N_VALID_ASMA, na.rm = TRUE),
      PCT_VALID_ASMA = N_VALID_ASMA / N_TOTAL,
      N_MISSING_RANGE = sum(N_MISSING_RANGE, na.rm = TRUE),
      N_MISSING_CLASS = sum(N_MISSING_CLASS, na.rm = TRUE),
      N_MISSING_RWY = sum(N_MISSING_RWY, na.rm = TRUE),
      N_MISSING_SECTOR = sum(N_MISSING_SECTOR, na.rm = TRUE),
      .by = YEAR
    ) |>
    mutate(PCT_VALID_ASMA = round(PCT_VALID_ASMA, 3)) |>
    kable()
}
```

```{r}
#| label: tbl-production-reference-coverage-check

if (file.exists(asma_reference_coverage_path)) {
  readr::read_csv(asma_reference_coverage_path, show_col_types = FALSE) |>
    summarise(
      N_REFERENCE_GROUPS = sum(N_REFERENCE_GROUPS, na.rm = TRUE),
      N_VALID_REFERENCE_GROUPS = sum(N_VALID_REFERENCE_GROUPS, na.rm = TRUE),
      .by = RANGE
    ) |>
    kable()
}
```
