---
title: "BRA - data preparation - network"
---

## Overview

There are several data sets to exploit for the network data portions.

* all movements
* ANAC
* punctuality data
* TOTALBR data (CGNA - Setamillenium)

## Data Loading

Punctuality data is currently stored in the BRA-EUR datalake.    
Download by hand and convert to parquet for reduced storage.    
Afterwards we "deactivate the code chunks by removing the r from {r}


```{}
# shortcut to download folder
pth_downloads <- here::here(here::here() |> dirname() |> dirname(), "Downloads")  

# movement data
mvts_fn <- fs::dir_ls(pth_downloads, regexp = "movement_data")
mvts <- readr::read_csv2(mvts_fn)
mvts |> arrow::write_parquet("./data-src/BRA-mvts-2023-2025.parquet")

# punctuality data
punc_fns <- fs::dir_ls(pth_downloads, regexp = "punctuality-BR")  # downloaded zip
archive  <- punc_fns |> basename()

# get out "raw data", helper functions rqutils-zip loaded outside this qmd
punc_raw_fns <- check_zip_content(pth_downloads, archive) |> 
  filter(grepl("-raw-", Name)) |> pull(Name)

punc_raw <- read_zip_content(pth_downloads, archive, punc_raw_fns, .type = "csv_auto") |> 
  bind_rows()
punc_raw |> arrow::write_parquet("./data-src/BRA-punc-raw.parquet")

# get other punc file - negate -raw-
punc_other_fn <- check_zip_content(pth_downloads, archive) |> 
  filter(! grepl("-raw-", Name)) |> pull(Name)
punc_other <- read_zip_content(pth_downloads, archive, punc_other_fn, .type = "csv_auto") 
```

punctuality0data-2025.csv is the combined and renamed raw-data files for 2025 with a row-number index.

let's do the same operation and write out as punc-2025.

```{}
punc_bra <- punc_raw |> 
    dplyr::rename(
       FLTID = Callsign
      , ADEP = `Sigla ADEP`, ADES = `Sigla ADES`
      , ICAO = `Sigla LOCALITY`
      , PHASE = Evento
      , SCHED_TIME = `DH Prev Calco Strat BRT`
      , BLOCK_TIME = `DH Real Calco Tat BRT`
      ) |> 
    dplyr::mutate(
        SCHED_TIME = lubridate::dmy_hms(SCHED_TIME)
      , BLOCK_TIME = lubridate::dmy_hms(BLOCK_TIME)
      )

punc_bra |> arrow::write_parquet("./data-src/BRA-punc-analytic-2025.parquet")
```

Get ANAC data - for 2025    
had to go quarter by quarter - apparently full year clashes with internet speed/access quota

```{}
# ensure latest package version
# install.packages("flightsbr")

month_codes <- sprintf(fmt = "%02d", seq(1,12, 1))   # 2 digits with leading zero
month_codes <- paste0("2025", month_codes)           # year-month

anac_2025_1 <- month_codes[1:3] |>                          # download data as loop
  purrr::map(.f = ~ flightsbr::read_flights(.x))
anac_2025_2 <- month_codes[4:6] |> 
  purrr::map(.f = ~ flightsbr::read_flights(.x))
# anac_2025_3 <- month_codes[7:9] |> 
#   purrr::map(.f = ~ flightsbr::read_flights(.x))
# anac_2025_4 <- month_codes[10:12] |> 
#   purrr::map(.f = ~ flightsbr::read_flights(.x))
# failed ... per month
#anac_202507 <- flightsbr::read_flights(month_codes[7])  # CRASHES - DATA FORMAT
anac_202508 <- flightsbr::read_flights(month_codes[8])
anac_202509 <- flightsbr::read_flights(month_codes[9])
#anac_202510 <- flightsbr::read_flights(month_codes[10]) # SERVER TIMEOUT/FAILURE
anac_202511 <- flightsbr::read_flights(month_codes[11])
anac_202512 <- flightsbr::read_flights(month_codes[12])

# glue everything together
anac_2025 <- list(
    anac_2025_1 |> bind_rows() |> as_tibble()
  , anac_2025_2 |> bind_rows() |> as_tibble()
  , anac_202508 |> as_tibble()
  , anac_202509 |> as_tibble()
  , anac_202511 |> as_tibble()
  , anac_202512 |> as_tibble() 
  ) |> 
  bind_rows()
  
# remove breadcrumbs
rm(anac_2025_1, anac_2025_2, anac_202508, anac_202509, anac_202511, anac_202512)
```

let's work with 2025 without the missing months

```{}
anac_2025 |> dplyr::rename(
   OPR_ICAO = sg_empresa_icao
  ,OPR_IATA = sg_empresa_iata
  ,FLT_NBR  = nr_voo
  ,DOF      = dt_referencia
  ,SVC      = cd_tipo_linha        # looks like service types - tbd
  ,TFC_TYPE = ds_natureza_etapa    # not sure about naming
  ,ADEP     = sg_icao_origem 
  ,ADEP_CNTRY = nm_pais_origem
  ,DEP_DAY  = dt_partida_real
  ,DEP_HR   = hr_partida_real
  , TYPE    = sg_equipamento_icao
  ,REG      = ds_matricula        # could be part of registration - tbd
  ,ADES     = sg_icao_destino 
  ,ADES_CNTRY = nm_pais_destino
  ,ARR_DAY  = dt_chegada_real
  ,ARR_HR   = hr_chegada_real
  
) |> 
  arrow::write_parquet("./data-src/BRA-ANAC-2025p.parquet")
```


read in ANAC data, trim to payload and apply standard naming (old maybe we can delete)

```{r}
anac_2025 <- arrow::read_parquet("./data-src/BRA-ANAC-2025p.parquet")

anac_2025 <- anac_2025 |> 
  dplyr::select(OPR_ICAO, FLT_NBR, TYPE, REG
                , ADEP, ADES
                , DEP_DAY, DEP_HR
                , ARR_DAY, ARR_HR 
                , ADEP_CNTRY, ADES_CNTRY
                , TFC_TYPE, SVC
                , DOF 
                , OPR_IATA
                ) |> 
  dplyr::mutate(
     FLTID = paste0(OPR_ICAO, FLT_NBR)
    , ATOT = lubridate::ymd_hms(paste(DEP_DAY, DEP_HR))
    , ALDT = lubridate::ymd_hms(paste(ARR_DAY, ARR_HR))
    ) |> 
  dplyr::select(FLTID, REG, TYPE, ADEP, ADES, ATOT, ALDT, everything())

anac_2025 |> arrow::write_parquet("./data-src/BRA-ANAC-2025p-trimmed.parquet")
```




### TotalBR

Read the file separate by year received from CGNA

```{}
# Step 1: Open the partitioned parquet dataset (year/month structure)
#ds <- arrow::open_dataset(here::here("data-src", "TOTALBR"))

# Step 2: Define the years to load
#target_years <- c(2023, 2024, 2025)

# Step 3: Filter by years and collect into a single tibble
#totalbr <- ds |>
#  dplyr::filter(year %in% target_years) |>
#  dplyr::collect()

# Step 4: Save the final dataset as a single parquet file
#arrow::write_parquet(totalbr, here::here("data-src", "totalbr.parquet"))
```




```{}
totalbr <- arrow::read_parquet(here::here("data-src", "totalbr.parquet"))

#unique_icaos <- c(bra_flt$ADEP, bra_flt$ADES) |> unique()

iso_meta <- readr::read_csv("./data/country-icao-iso-etc.csv", show_col_types = F) |> 
  dplyr::rename(CNTRY_ISO = iso2c, CNTRY_NAME = country.name.en, ICAO_REGION = icao.region)

oa <- read_csv(here::here("data","oa-202603.csv"), show_col_types = FALSE) |> 
  filter(!is.na(icao_code)) |> 
  select(ICAO = icao_code, CNTRY_ISO = iso_country)

# path oa for Brazil
patch_oa <- tibble::tribble(
  ~ ICAO, ~ CNTRY_ISO
  ,"SBCO", "BR"
  ,"SBJD",  "BR"    
  ,"SBJI",  "BR"
  ,"SBMI",  "BR"
 ,"SBMT" ,  "BR"
 ,"SBPO",  "BR"
 ,"SBRD" , "BR"
 ,"SBSG" , "BR"
 ,"SBSI",  "BR"
 ,"SBSO",  "BR"
,"SBUY",  "BR"
,"SBVC" , "BR"
,'GQNN',  "MR"    # Maritania
,'SDAG', 'BR'
,'SDIY', 'BR'
,'SDJV', 'BR'
,'SDLO', 'BR'
,'SDVZ', 'BR'
,'SIFC', 'BR'
,'SIRI', 'BR'
,'SIUA', 'BR'
,'SIXE', 'BR'
,'SJZA', 'BR'
,'SN6L', 'BR'
,'SNAB', 'BR'
,'SNBA', 'BR'
,'SNCL', 'BR'
,"SNEB",  "BR"
,"SNIG" , "BR"
,"SNJM"  , "BR"
,"SNLN" , "BR"
,"SNRJ"  , "BR"
,"SNSM"  ,  "BR"
,"SNSS"  ,  "BR"
,"SNTS"  ,  "BR"
,"SNWS"  ,  "BR"
,"SNZR"  ,  "BR"
,"SSBN"  ,  "BR"
,"SSGG"  ,  "BR"
,"SSQO"  ,  "BR"
,"SSRS"  ,  "BR"
,"SSVL"  ,  "BR"
,"SWBE"  ,  "BR"
,"SWKQ"  ,  "BR"
,"SWYN"  ,  "BR"
,"SYEC"  ,  "GN"
, "SADF" ,  "AG" #argentina
, "FOAR" , "ZA" # South africa
)
oa <- oa |> dplyr::bind_rows(patch_oa)

ndf <- totalbr |> dplyr::select(
    FLTID  = co_indicativo,
    ADEP   = co_addep,
    ADES   = co_addes,
    TYPE   = co_modelo,
    DATE   = dt_dia,
    SVC    = li_tipovoo
) 

#oa <- read_csv(here::here("data","oa-202603.csv")) |> 
 # filter(!is.na(icao_code)) |> 
  #select(ICAO = icao_code, CNTRY = iso_country)

ndf2 <- ndf |> 
  left_join(oa, join_by(ADEP == ICAO)) |> 
  rename(ADEP_CNTRY = CNTRY_ISO) |> 
  left_join(oa, join_by(ADES == ICAO)) |> 
  rename(ADES_CNTRY = CNTRY_ISO) 
 
ndf3 <- ndf2 |> 
    mutate(
        ADEP_CNTRY = case_when(
            grepl("^S[BDNSWISJ]|9|^Z|AFIL|NI", ADEP) & is.na(ADEP_CNTRY) ~ "BR"
            ,.default = ADEP_CNTRY
        )
        , ADES_CNTRY = case_when(
            grepl("^S[BDNSWISJ]|9|^Z|AFIL|NI", ADES) & is.na(ADES_CNTRY) ~ "BR"
            ,.default = ADES_CNTRY
        )
    ) |> 
    mutate(DAIO = case_when(
         ADEP_CNTRY == "BR" & ADES_CNTRY == "BR" ~ "I"
        ,ADEP_CNTRY != "BR" & ADES_CNTRY != "BR" ~ "O"
        ,ADEP_CNTRY == "BR" & ADES_CNTRY != "BR" ~ "D"
        ,ADEP_CNTRY != "BR" & ADES_CNTRY == "BR" ~ "A"
        ,.default = NA
    )
  )


 

```

Traffic overview

```{r}
#| message: false

magic_number <- 1.28

bra_tfc <- dplyr::bind_rows(
   readr::read_csv("./data/BRA-network-tfc-movts-all-RafHug.csv") |> mutate(SOURCE = "RafHug")
  ,readr::read_csv("./data/BRA-network-tfc-2025.csv") |> 
    mutate(SOURCE = "Bene") |> 
    rename(DLY_FLTS = N, MVTS_NORM_ROLLAVG = N_NORM) |> 
    mutate(DLY_FLTS = magic_number * DLY_FLTS
           ,MVTS_NORM_ROLLAVG = magic_number * MVTS_NORM_ROLLAVG)
)
```

```{r}
bra_tfc |> 
    ggplot(aes(x = DATE)) +
    geom_path(aes(y = MVTS_NORM_ROLLAVG, group = SOURCE, color = SOURCE) ) +    #, colour = bra_col) +
    geom_point(aes(y = DLY_FLTS, color = SOURCE) #, colour = bra_col
               , alpha = 0.2, size = 0.2)
```

aerodrome ranking based on departures

create lookup table based on NM-flt

```{r}
# get some of the rq helper functions to read NM flight table - only works on RQ's machine
# path-to-flight-table
pth_nm <- here::here( here::here() |> dirname(), "__DATA", "NM-flight-table")
nm_flt <- read_zip_content(pth_nm, "NM-flt-2025.zip")
```

```{r}
patch_unknown_icao_location_indicators <- function(
      df_with_iso2c
    , icao_var  = ICAO  
    , iso2c_var = CNTRY_ISO
){
  df_patched <- df_with_iso2c |> 
  #------- patch unknown airports ------------------
  dplyr::mutate({{ iso2c_var  }} := dplyr::case_when(
     grepl(pattern = "^EB", {{ icao_var }}) & is.na({{ iso2c_var}}) ~ "BE"
    ,grepl(pattern = "^E[DT]", {{ icao_var }}) & is.na({{ iso2c_var}}) ~ "DE"
    ,grepl(pattern = "^EF", {{ icao_var }}) & is.na({{ iso2c_var}}) ~ "FI"
    ,grepl(pattern = "^EG", {{ icao_var }}) & is.na({{ iso2c_var}}) ~ "GB"
    ,grepl(pattern = "^EH", {{ icao_var }}) & is.na({{ iso2c_var}}) ~ "NL"
    ,grepl(pattern = "^EI", {{ icao_var }}) & is.na({{ iso2c_var}}) ~ "IE"
    ,grepl(pattern = "^EN", {{ icao_var }}) & is.na({{ iso2c_var}}) ~ "NO"
    ,grepl(pattern = "^EK", {{ icao_var }}) & is.na({{ iso2c_var}}) ~ "DK"
    ,grepl(pattern = "^EP", {{ icao_var }}) & is.na({{ iso2c_var}}) ~ "PL"
    ,grepl(pattern = "^ES", {{ icao_var }}) & is.na({{ iso2c_var}}) ~ "SE"
    ,grepl(pattern = "^LE", {{ icao_var }}) & is.na({{ iso2c_var}}) ~ "ES"
    ,grepl(pattern = "^LF", {{ icao_var }}) & is.na({{ iso2c_var}}) ~ "FR"
    ,grepl(pattern = "^LG", {{ icao_var }}) & is.na({{ iso2c_var}}) ~ "GR"
    ,grepl(pattern = "^LH", {{ icao_var }}) & is.na({{ iso2c_var}}) ~ "HU"
    ,grepl(pattern = "^LI", {{ icao_var }}) & is.na({{ iso2c_var}}) ~ "IT"
    ,grepl(pattern = "^LK", {{ icao_var }}) & is.na({{ iso2c_var}}) ~ "CZ"
    ,grepl(pattern = "^LL", {{ icao_var }}) & is.na({{ iso2c_var}}) ~ "IL"    # Israel
    ,grepl(pattern = "^LO", {{ icao_var }}) & is.na({{ iso2c_var}}) ~ "AT"
    ,grepl(pattern = "^LP", {{ icao_var }}) & is.na({{ iso2c_var}}) ~ "PT"
    ,grepl(pattern = "^LR", {{ icao_var }}) & is.na({{ iso2c_var}}) ~ "RO"
    ,grepl(pattern = "^LS", {{ icao_var }}) & is.na({{ iso2c_var}}) ~ "CH"
    ,grepl(pattern = "^GM", {{ icao_var }}) & is.na({{ iso2c_var}}) ~ "MA"    # Morocco
    #--------------- international / non-ECTRL
    # RQ "old knowledge"  or https://en.wikipedia.org/wiki/Lists_of_airports#By_code
    ,grepl(pattern = "^FY", {{ icao_var }}) & is.na({{ iso2c_var}}) ~ "NA"    # Namibia
    ,grepl(pattern = "^HE", {{ icao_var }}) & is.na({{ iso2c_var}}) ~ "EG"    # Egypt
    ,grepl(pattern = "^K", {{ icao_var }}) & is.na({{ iso2c_var}}) ~ "US" # USA -> KL?
    ,grepl(pattern = "^OK", {{ icao_var }}) & is.na({{ iso2c_var}}) ~ "KW" # Brazil 
    ,grepl(pattern = "^UD", {{ icao_var }}) & is.na({{ iso2c_var}}) ~ "AM" # Armenia
    ,grepl(pattern = "^UT", {{ icao_var }}) & is.na({{ iso2c_var}}) ~ "TJ" # Tajikistan, Turkmenistan
    ,grepl(pattern = "^SB", {{ icao_var }}) & is.na({{ iso2c_var}}) ~ "BR" # Brazil 
    ,.default = {{ iso2c_var }}
  )) 
  return(df_patched)
}
```




Test regions

```{r}

tmp <- ndf3 |> left_join(isometa2 , join_by(ADEP_CNTRY == CNTRY_ISO)) |> 
  rename( DEPREG = eurocontrol_pru) |> 
  left_join(isometa2 , join_by(ADES_CNTRY == CNTRY_ISO)) |> 
  rename( ARRREG = eurocontrol_pru)





```


# DATA PREPARATION for traffic graph since 2019 Chapter 3


```{r}

# Brazil data

# totalbr <- arrow::read_parquet(here::here("data-src", "totalbr.parquet")) #local file 1gb
# 
# ndf <- totalbr |> dplyr::select(
#     FLTID  = co_indicativo,
#     ADEP   = co_addep,
#     ADES   = co_addes,
#     TYPE   = co_modelo,
#     DATE   = dt_dia,
#     SVC    = li_tipovoo
# )

# ndf |> arrow::write_parquet(here::here("data-src" , "ndf_bra_totalbr.parquet"))

## -------  BRA standart data stopped here ---------

tmp_bra <- ndf |> mutate(DATE = as.Date(DATE)) |>
    dplyr::group_by(DATE) |>
  dplyr::reframe(DLY_FLTS = n()) |>
  dplyr::mutate(
  MVTS_NORM_ROLLAVG = zoo::rollmedian(DLY_FLTS, k = 7, na.pad=TRUE))
 
tmp_bra |> arrow::write_parquet(here::here("data" , "ndf_bra_totalbr_daily.parquet"))


## -------  BRA standart data stopped here ---------

## Auxiliary function to plot. 

## ------ Lines below should be move to report -----

tmp_bra <- arrow::read_parquet(here::here("data-src", "ndf_bra_totalbr_daily.parquet"))

plot_region <- function(data, color_raw, color_avg, title_label) {
  ggplot2::ggplot(data, aes(x = DATE)) +
    geom_point(
      aes(y = DLY_FLTS),
      color  = color_raw,
      alpha  = 0.2,
      size   = 0.3
    ) +
    geom_path(
      aes(y = MVTS_NORM_ROLLAVG),
      color     = color_avg,
      linewidth = 0.7
    ) +
    scale_y_continuous(
      limits = c(0, NA),          
      expand = expansion(mult = c(0, 0.05))
    ) +
    scale_x_date(
      date_breaks = "2 years",
      date_labels = "%Y"
    ) +
    labs(
      title = title_label,
      x     = NULL,
      y     = NULL
    ) +
    theme_minimal(base_size = 11) +
    theme(
      plot.title   = element_text(hjust = 0.5, size = 11),
      panel.grid.minor = element_blank()
    )
}

# --- Panels ---
p_brazil <- plot_region(
  data       = tmp_bra,                  
  color_raw  = "#99cc99",             
  color_avg  = "#2e7d32",            
  title_label = "Brazil Region daily movement (rolling 7-day average)"
)

p_europe <- plot_region(
  data       = tmp_eur,              
  color_raw  = "#90caf9",            # azul claro (pontos)
  color_avg  = "#1565c0",            # azul escuro (linha)
  title_label = "European Region daily movement (rolling 7-day average)"
)

# --- Combine ---
p_brazil / p_europe


```


# Figure 3.2 Comparing traffic


```{r}




# Brazilian data

# tmp_bra_yoy <- ndf |>
#   dplyr::mutate(DATE = as.Date(DATE)) |>
#   dplyr::filter(
#     lubridate::year(DATE) %in% c(2023, 2024, 2025)
#   ) |>
#   dplyr::group_by(DATE) |>
#   dplyr::reframe(DLY_FLTS = n()) |>
#   dplyr::mutate(
#     YEAR = factor(lubridate::year(DATE)),
#     DOY  = lubridate::yday(DATE)
#   ) |>
#   dplyr::group_by(YEAR) |>
#   dplyr::mutate(
#     MVTS_ROLLAVG = zoo::rollapply(
#       DLY_FLTS, width = 28, FUN = median,
#       partial = TRUE, align = "center"
#     )
#   ) |>
#   dplyr::ungroup()
# 
# tmp_bra_yoy |> arrow::write_parquet(here::here("data" , "ndf_bra_totalbr_annually.parquet"))

# European data


## -------  BRA standart data stopped here ---------

## ------ Lines below should be move to report -----

tmp_bra_yoy <- arrow::read_parquet(here::here("data-src", "ndf_bra_totalbr_annually.parquet"))

YEAR_COLORS <- c("2023" = "#E74C3C",   
                 "2024" = "#2ECC71",   
                 "2025" = "#5DADE2")   

plot_region_yoy <- function(data, title_label) {
  ggplot2::ggplot(data, ggplot2::aes(x = DOY, colour = YEAR)) +
      ggplot2::geom_line(
      ggplot2::aes(y = MVTS_ROLLAVG),
      linewidth = 0.7,
      na.rm     = TRUE        
) +
    ggplot2::scale_colour_manual(values = YEAR_COLORS) +
    ggplot2::scale_x_continuous(
          breaks = c(1, 32, 60, 91, 121, 152, 182, 213, 244, 274, 305, 335),
          labels = c("Jan","Feb","Mar","Apr","May","Jun",
                      "Jul","Aug","Sep","Oct","Nov","Dec")
    ) +
    ggplot2::scale_y_continuous(
      limits = c(0, NA),
      expand = ggplot2::expansion(mult = c(0, 0.05))
    ) +
    ggplot2::labs(
      title  = title_label,
      x      = NULL,
      y      = NULL,
      colour = NULL
    ) +
    ggplot2::theme_minimal(base_size = 11) +
    ggplot2::theme(
      plot.title       = ggplot2::element_text(hjust = 0.5, size = 11),
      panel.grid.minor = ggplot2::element_blank(),
      legend.position  = "bottom"
    )
}

# --- Panels ---

p_brazil_yoy <- plot_region_yoy(tmp_bra_yoy, "Brazil Region – Daily Movements by Year")


p_europe_yoy <- plot_region_yoy(tmp_eur_yoy, "European Region – Daily Movements by Year")



p_brazil_yoy / p_europe_yoy


```






