# Brazil Taxi Data Ingestion

```{r}
source(here::here("_chapter-setup.R"))
```

This note documents the preparation of the Brazilian taxi-time files used for the 2025 Brazil-Europe operational efficiency update.
The raw taxi files were downloaded from the DECEA datalake folder on OneDrive/SharePoint (`Datalake-Rafael/decea-2025/TAXI`) to the local Downloads folder as a transient source archive.
To save local storage, the raw CSV files and the raw zip archive are not copied into the project archive.

The retained archive artefacts are filtered, study-airport extracts in an APDF-like parquet format, stored in:

```{r}
bra_apdf_dir <- "/Users/rainerkoelle/RProjects/PBWG-BRA/data/apdf"
bra_output_dir <- "/Users/rainerkoelle/RProjects/PBWG-BRA/outputs"
bra_report_data <- here::here("data")
```

## Source Inventory

```{r}
#| label: tbl-bra-source-inventory

zip_path <- "/Users/rainerkoelle/Downloads/OneDrive_1_20-04-2026.zip"

if (file.exists(zip_path)) {
  utils::unzip(zip_path, list = TRUE) |>
    dplyr::filter(grepl("^dsTaxi20[0-9]{2}\\.csv$", Name)) |>
    dplyr::mutate(Size_MB = round(Length / 1024^2, 1)) |>
    dplyr::select(Name, Size_MB, Date) |>
    knitr::kable()
} else {
  tibble::tibble(
    Name = c("dsTaxi2023.csv", "dsTaxi2024.csv", "dsTaxi2025.csv"),
    Note = "Raw source archive is not retained locally in the project archive."
  ) |>
    knitr::kable()
}
```

The source files use semicolon-separated fields and include Brazilian airport movements beyond the Brazil-Europe study scope.
The harmonised parquet extracts retain only the airports defined in `_chapter-setup.R`.

## Harmonisation

The payload columns are mapped to the APDF-like fields required by the taxi-time pipeline.
For arrivals, the airport is derived from `addestino`; for departures, the airport is derived from `adpartida`.

| Source field | Harmonised field | Note |
| --- | --- | --- |
| `indicativo` | `FLTID` | Flight identifier |
| `matricula` | `REG` | Aircraft registration |
| `adpartida` | `ADEP` | Departure aerodrome |
| `addestino` | `ADES` | Destination aerodrome |
| `mov` | `PHASE` | `Arr` to `ARR`, `Dep` to `DEP` |
| `dh_bimtra` | `MVT_TIME` | Movement time |
| `dh_vra` | `BLOCK_TIME` | Block time |
| `tipoaeronave` | `ARCTYP` | Aircraft type |
| `box` | `STND` | Stand or gate position |
| `pista` | `RWY` | Runway |
| `match_vra` | `MATCH_VRA` | Payload match flag |
| trailing field when present | `KPI_PAYLOAD` | Retained for later diagnostic checks, not used in the harmonised metric |

The 2024 source file contains a trailing payload field on many rows.
The ingestion therefore reads up to 15 semicolon-separated fields and keeps the additional value as `KPI_PAYLOAD`.
This prevents a column shift in the 2024 data while leaving the supplied KPI calculation out of the PBWG taxi-time metric.

## Preparation Code

The following chunk is the reproducible preparation script used to create the filtered parquet extracts and the daily taxi-time outputs.
It is not evaluated during normal rendering because it reads the large transient source archive from Downloads.

```{r}
#| label: prepare-bra-taxi-data
#| eval: false

library(dplyr)
library(readr)
library(lubridate)
library(purrr)
library(fs)
library(arrow)
library(data.table)

devtools::load_all("/Users/rainerkoelle/RProjects/PBWG", quiet = TRUE)
source("/Users/rainerkoelle/RProjects/BRA-EUR-2025/_chapter-setup.R")

zip_path <- "/Users/rainerkoelle/Downloads/OneDrive_1_20-04-2026.zip"
archive_root <- "/Users/rainerkoelle/RProjects/PBWG-BRA"
report_data_dir <- "/Users/rainerkoelle/RProjects/BRA-EUR-2025/data"
variant <- "icao_ganp_p20"
ref_year <- 2024L
min_n <- 5L
max_txxt <- 120
study_airports <- bra_apts

dir_create(path(archive_root, "data", "apdf"))
dir_create(path(archive_root, "outputs", "txxt-reference-2024"))
walk(2023:2025, ~ dir_create(path(archive_root, "outputs", paste0("txxt-daily-", .x))))

normalise_text <- function(x) {
  x <- as.character(x)
  x <- trimws(x)
  x[x == ""] <- NA_character_
  x
}

parse_bra_datetime <- function(x) {
  x <- normalise_text(x)
  parse_date_time(x, orders = c("ymd HMS", "ymd HMS OS"), tz = "UTC", quiet = TRUE)
}

read_bra_taxi_year <- function(year) {
  member <- paste0("dsTaxi", year, ".csv")
  message("Reading ", member)

  raw <- fread(
    cmd = sprintf("unzip -p %s %s", shQuote(zip_path), shQuote(member)),
    sep = ";",
    fill = 15,
    colClasses = "character",
    showProgress = FALSE
  ) |>
    as_tibble()

  if (!"V15" %in% names(raw)) {
    raw$V15 <- NA_character_
  }

  harmonised <- raw |>
    transmute(
      FLTID = normalise_text(indicativo),
      REG = normalise_text(matricula),
      ADEP = normalise_text(adpartida),
      ADES = normalise_text(addestino),
      PHASE = recode(normalise_text(mov), Arr = "ARR", Dep = "DEP", .default = NA_character_),
      MVT_TIME = parse_bra_datetime(dh_bimtra),
      BLOCK_TIME = parse_bra_datetime(dh_vra),
      ARCTYP = normalise_text(tipoaeronave),
      STND = normalise_text(box),
      RWY = normalise_text(pista),
      MATCH_VRA = normalise_text(match_vra),
      KPI_PAYLOAD = normalise_text(V15),
      ICAO = case_when(
        PHASE == "ARR" ~ ADES,
        PHASE == "DEP" ~ ADEP,
        TRUE ~ NA_character_
      )
    ) |>
    filter(ICAO %in% study_airports)

  write_parquet(
    harmonised,
    path(archive_root, "data", "apdf", paste0("PBWG-BRA-dsTaxi-apdf-", year, ".parquet"))
  )

  harmonised
}

samples_by_year <- map(2023:2025, function(year) {
  read_bra_taxi_year(year) |>
    prepare_txxt_reference_input(max_txxt = max_txxt) |>
    mutate(YEAR = year)
}) |>
  set_names(as.character(2023:2025))

reference_samples <- samples_by_year[[as.character(ref_year)]]
ref_start <- ymd_hms(paste0(ref_year, "-01-01 00:00:00"), tz = "UTC")
ref_end <- ymd_hms(paste0(ref_year, "-12-31 23:59:59"), tz = "UTC")

reference_data <- build_txxt_reference(
  txxt_samples = reference_samples,
  ref_start = ref_start,
  ref_end = ref_end,
  variant = variant,
  min_n = min_n,
  keep_below_threshold = TRUE,
  include_unknown = FALSE
)

walk(unique(reference_data$ICAO), function(airport) {
  write_pbwg_txxt_reference(
    data = reference_data,
    airport = airport,
    ref_period = as.character(ref_year),
    variant = variant,
    min_n = min_n,
    output_dir = path(archive_root, "outputs", "txxt-reference-2024"),
    region = "BRA"
  )
})

daily_by_year <- imap(samples_by_year, function(samples, year_chr) {
  year <- as.integer(year_chr)
  message("Applying 2024 reference to ", year)

  augmented <- apply_txxt_reference(
    txxt_samples = samples,
    reference_data = reference_data,
    valid_reference_only = TRUE
  )

  daily <- summarise_pbwg_txxt_daily(augmented, year = year)
  daily_dir <- path(archive_root, "outputs", paste0("txxt-daily-", year))

  walk(unique(daily$ICAO), function(airport) {
    write_pbwg_txxt(
      data = daily,
      year = year,
      ref_period = ref_year,
      variant = variant,
      output_dir = daily_dir,
      airport = airport,
      region = "BRA"
    )
  })

  combined_path <- write_pbwg_txxt(
    data = daily,
    year = year,
    ref_period = ref_year,
    variant = variant,
    output_dir = daily_dir,
    airport = NULL,
    region = "BRA"
  )

  file_copy(combined_path, path(report_data_dir, basename(combined_path)), overwrite = TRUE)
  daily
})
```

## Retained Artefacts

```{r}
#| label: tbl-bra-retained-artefacts

parquet_files <- list.files(bra_apdf_dir, pattern = "\\.parquet$", full.names = TRUE)

tibble::tibble(path = parquet_files) |>
  dplyr::mutate(
    file = basename(path),
    size_mb = round(file.info(path)$size / 1024^2, 1)
  ) |>
  dplyr::select(file, size_mb) |>
  knitr::kable()
```

The daily analytic outputs copied into the report project are:

```{r}
#| label: tbl-bra-report-outputs

report_output_files <- list.files(
  bra_report_data,
  pattern = "^PBWG-BRA-txxt-analytic-20[0-9]{2}-ref2024-icao_ganp_p20\\.csv$",
  full.names = FALSE
)

tibble::tibble(file = report_output_files) |>
  knitr::kable()
```

## Coverage Check

The pipeline records the number of valid taxi-time movements (`MVTS_VALID`) and the number of movements that could not be used for the metric (`MVTS_NA`).
The observed taxi-time total (`TOT_TXXT`), reference total (`TOT_REF`), and additional-time total (`TOT_ADD_TIME`) are calculated only over `MVTS_VALID`.

```{r}
#| label: tbl-bra-txxt-coverage

readr::read_csv(
  here::here("data", "BRA-txxt-coverage-summary-2023-2025.csv"),
  show_col_types = FALSE
) |>
  dplyr::mutate(
    MVTS_NA_SHARE = MVTS_NA / MVTS_TOTAL,
    AVG_TXXT = TOT_TXXT / MVTS_VALID,
    AVG_ADD_TIME = TOT_ADD_TIME / MVTS_VALID
  ) |>
  dplyr::select(YEAR, PHASE, MVTS_TOTAL, MVTS_VALID, MVTS_NA, MVTS_NA_SHARE, AVG_TXXT, AVG_ADD_TIME) |>
  dplyr::mutate(
    dplyr::across(c(MVTS_NA_SHARE, AVG_TXXT, AVG_ADD_TIME), ~ round(.x, 3))
  ) |>
  knitr::kable()
```

The current extracts retain the source KPI payload in the harmonised parquet files for later diagnostic comparison.
It is not used in the 2024 GANP 20th percentile PBWG reference calculation.
